---
title: "BLPW_assignment"
author: "Sydney_Bliss"
date: '2018-02-21'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### ASSIGNMENT - BLACKPOLL WARBLER RELATIVE MASS PLOT  

**STEP 1 - import packages, look at data**

```{r results='hide', message=F, warning=F}
library(tidyverse) 
library(lubridate)
rm(list = ls())
```

Open the RDS file, then View to see the data:
```{r}
blpw.all <- readRDS("blpw.all.RDS")
```

Check what variables are in each column:
```{r}
table(blpw.all$location) # 2 options: BP (n=1892) and SI (n=566)
table(blpw.all$age) # 5 options: AHY (n=523), ASY (n=0), HY (n=1802), SY (n=0), and U (n=133)
table(blpw.all$recap) # 2 options: N (n=2163), R (n=295), U (n=0)
table(blpw.all$year) # 16 options: 1996-2008, 2010-2012
table(blpw.all$month) # 3 options: Aug, Sep, Oct 
```

**STEP 2 - manipulate data**  

Find band numbers with multiple records (i.e. recaps) and isolate them:
```{r}
blpw.all <- blpw.all %>%
            group_by(band) %>% 
            filter(n() > 1) 

blpw.all # manipulation looks ok
```

Set dates to same year so time stamp reads them properly, and recode dates as a single variable:
```{r}
blpw.all <- blpw.all %>%
            mutate(year, year = 2000) %>% 
            mutate(date = make_date(year, month, day))

blpw.all # manipulation looks ok
```

See how many bands are recaps, and # of recaps for each band (not necessary for analysis, but useful):
```{r}
blpw.recap.sums <- blpw.all %>%
                   group_by(band, location) %>%
                   summarize(n())

blpw.recap.sums # manipulation looks ok
```

**STEP 3 - get within-band mass differences**
```{r}
blpw.all <- blpw.all %>% 
            group_by(band) %>% 
            mutate(rel.mass = mass - first(mass))

blpw.all # manipulation looks ok
```

**STEP 4 - plot data**

Can't get colour to work, so leave as black and white:
```{r}
ggplot(blpw.all, aes(x = date, y = rel.mass, group = band)) +
  geom_line()  + geom_point() +
  xlab("Time of Year") + ylab("Mass (in grams, relative to capture date)")  +
  facet_grid(. ~ location) 
```



