---
title: "R_Markdown"
author: "Sydney_Bliss"
date: '2018-01-31'
output: html_document
---
 
#####Swallow Breeding Performance & Phenology in the Maritimes  

**Background**
Aerial insectivore populations, including Tree (Tachycineta bicolor) and Barn (Hirundo rustica) Swallows, have experienced precipitous declines since the 1970s in North America. There are indications that there may be a concurrent decline in insect populations, coupled with earlier spring emergence driven by climate change. Such a decline in insect populations would likely alter the breeding phenology and performance of swallows, and may be linked to species-wide population declines. In the absence of insect abundance data, weather data - which influences insect abundance - can instead be used to examine relationships between food availability, and swallow breeding phenology and performance.

**Biological Questions:**   
Q1 - Has swallow breeding phenology (clutch initiation date) changed from 1962-2016?    
Hypothesis: both spp. will have advanced CID over time  
Rationale: insects are likely emerging earlier due to advanced spring thaw-up, and swallows have adjusted their phenology in response  

Q2 - Has swallow breeding performance (chick survival) changed from 1962-2016? 
Hypothesis: chick survival has decreased over time, contributing to the overall population declines in both spp.  
Rationale: decreased chick survial (one measure of breeding performance) would be an obvious explanation as to why populations have broadly declined. If chick survial has not changed, it suggests declines may be occuring at other points of the annual cycle (ex. high mortality during migration or overwintering)   
NOTE: could also examine other measures of breeding performance such as clutch size and brood size  

Q3 - Is breeding phenology affected by climate?   
Hypothesis: rainfall and temperature variables will explain the most variation in clutch initiation date    Rationale: rainfall and temperature are both positively associated with insect abundance, and likely emergence    

Q4 - Is breeding performance affected by climate?    
Hypothesis: chick survival has been negatively affected by changes in climate  
Rationale: if swallow breeding performance has declined over time, the most obvious explanation is a decrease in insect populations driven by climate change
  
**STEP 1: IMPORT PACKAGES & DATA**  

Load required packages, import data set and save it to an object:
```{r chunk_namee, results='hide'}
library(tidyverse) 
rm(list = ls())
swallows.df <- read_csv("swallows.csv")
```

Make sure R is reading the data correctly:
```{r}
problems(swallows.df) 
```
**STEP 2: TIDY THE DATA**   

Drop unnecessary columns using select():
```{r}
swallows.df <- select(swallows.df, Species, NestID, Province, Site, Latitude, Longitude, Year, CID, HD, CS, BS, SD12, minFt, minJFt, medJFt, medJp, meanMAp)
```

Create three time bins, (1) 1960-1972, (2) 1973-2005, (3) 2006-2016 to aid interpretation of changes over time, using mutate():
```{r}
swallows.df <- swallows.df %>% mutate(Yr = cut(Year, breaks = c(1959, 1972, 2005, 2016), labels =  c("1960-1972", "1973-2005", "2006-2016")))
```

Create survival rate (Surv.Rate) by dividing the SD12 variable (# of chicks that survived to day 12) by clutch size (CS), using mutate():
```{r}
swallows.df <- swallows.df %>% mutate(Surv.Rate = SD12 / CS)
```

There’s an outlier in a Tree Swallow CID record (value = 96, or Aug 4) which is biologically unlikely. Probably due to a recording error. Drop outlier:
```{r}
swallows.df <- filter(swallows.df, CID < 96) 
```

Make sure the data look ok after all these changes:
```{r}
summary(swallows.df)
str(swallows.df)
```

Make a tibble for each species so they can be examined separately:
```{r}
TRES.df <- filter(swallows.df, Species == "TRES")  # tree swallows
BARS.df <- filter(swallows.df, Species == "BARS")  # barn swallows
```
   
**STEP 3: VISUALIZE DATA for each biological question**

Histogram to see how many records exist for each species. Result looks unbalanced, which is likely due to varying volunteer effort. Far more Tree Swallows were sampled than Barn Swallows, and most data were collected after 2010 in both species.
```{r}
ggplot(swallows.df, aes(x = Year)) + 
  geom_histogram(bins = 60) + 
  facet_grid(. ~ Species) + 
  xlab("Year") + 
  ylab("Count") +
  labs(title = "Number of Nests Sampled, Tree and Barn Swallows")
```
   
**Q1 - Changes in breeding phenology. Look at clutch initiation date (CID) in both species over time.**

Scatterplot to see spread of the data. Looks like Barn Swallows start laying eggs later in the spring and continue later into the summer than Tree Swallows, likely because Barn Swallows are double-brooded. Tree Swallows start laying earlier and have a more restricted laying range.
```{r}
ggplot(swallows.df, aes(Year, CID)) + 
  geom_point() + 
  geom_jitter() +
  xlab("Year") + 
  ylab("Clutch initiation date (May 1 = day 1)") + 
  facet_grid(. ~ Species)
```

Boxplot to see the same data another way. Looks like the both species are initiating clutches earlier in the 2006-16 period than they were in the 1960-72 period.
```{r}
ggplot(swallows.df, aes(Yr, CID)) + 
  stat_boxplot(geom="errorbar") + 
  geom_boxplot(outlier.shape=1) + 
  xlab("Period") + ylab("Clutch initiation date (May 1 = day 1)") + 
  facet_grid(. ~ Species)
```
   
**Q2 - Changes breeding performance. Look at % of the clutch that survived (Surv.Rate) in both species over time.**  

The boxplot and scatterplot are difficult to interpret alone and more informative when combined. It looks like survival is fairly consistent over time. Perhaps it has decreased slightly in Barn Swallows.
```{r chunk_name, warning=FALSE}
ggplot(swallows.df, aes(Yr, Surv.Rate)) + 
  stat_boxplot(geom="errorbar") + 
  geom_boxplot(outlier.shape=1) + 
  xlab("Period") + ylab("Survival") + 
  facet_grid(. ~ Species)  +   
  geom_jitter()
```
Try with alpha:

```{r}
ggplot(swallows.df, aes(Yr, Surv.Rate)) +   
  stat_boxplot(geom="errorbar") + 
  geom_boxplot(outlier.shape=1) + 
  xlab("Period") + ylab("Survival") + 
  facet_grid(. ~ Species)  +
  geom_point(alpha=0.3)
```


**Q3 - Breeding phenology and climate variables. Is there a relationship between clutch initiation date (CID) and various measures on climate?**

Note the BARS clutches are not seperated here, limiting the interpretation for this species.

Climate variable #1 - minimum February temperature (minFt). Looks like minFt has a strong negative relationship with CID:

```{r}
ggplot(swallows.df, aes(minFt, CID)) + 
    geom_smooth(mapping = aes(x = minFt, y = CID)) + 
    xlab("Minimum February temperature (C)") + 
    ylab("Clutch initiation date (May 1 = day 1)") + 
    facet_grid(. ~ Species)
```

Climate variable #2 - minimum January-February temperature (minJFt). Looks like minJFt has a negative relationship with CID in TRES, but not in BARS:

```{r}
  ggplot(swallows.df, aes(minJFt, CID)) + 
    geom_smooth(mapping = aes(x = minJFt, y = CID)) + 
    xlab("Minimum Jan-Feb temperature (C)") + 
    ylab("Clutch initiation date (May 1 = day 1)") + 
    facet_grid(. ~ Species)
```

Climate variable #3 - median January-February temperature (medJFt). No consistent relationship in either species:
```{r}
ggplot(swallows.df, aes(medJFt, CID)) + 
  geom_smooth(mapping = aes(x = medJFt, y = CID)) + 
  xlab("Median Jan-Feb temperature (C)") + 
  ylab("Clutch initiation date (May 1 = day 1)") + 
  facet_grid(. ~ Species)
```

Climate variable #4 - mean annual precipitation (meanMAp). Clutch initiation date is  earliest with meanMAp is lowest and almost highest. Does this make sense?

```{r}
ggplot(swallows.df, aes(meanMAp, CID)) + 
  geom_smooth(mapping = aes(x = meanMAp, y = CID)) + 
  xlab("Mean monthly precipitation (cm)") + 
  ylab("Clutch initiation date (May 1 = day 1)") + 
  facet_grid(. ~ Species)
```
